\usefont{OT1}{bch}{b}{n} \normalfont \normalsize \textsc{STEM Fellowship Big Data Challenge 2020} \\ [10pt] \huge Deep learning transcriptomic model for prediction of pan-drug chemotherapeutic sensitivity \\

### 2020-06-01 manuscript/manuscript.tex
final copy

% Abstract
\begin{abstract}
	Emerging precision oncology studies have yet to generate a predictive biomarker that utilizes gene expression profiles to stratify tumours into similar pan-drug sensitivity profiles. This development would allow for identification of candidate drugs for treatments that maximize therapeutic response and minimize cytotoxic burden. As such, this study utilized cell line sensitivity and molecular profiling data to generate a combinatorial gene expression predictive biomarker, utilizing feature selection and a deep learning model. A pan-cancer cohort of cell-line gene expression data from Genomics of Drug Sensitivity in Cancer (GDSC) was clustered into two response groups. Cell line response groups showed a significant difference in pan-drug chemotherapeutic sensitivity. Due to the high dimensional nature of the microarray data, biologically agnostic feature selection was conducted to identify genes with the highest predictive value. The feature space was reduced to 300 genes, which functional profiling indicated was primarily enriched for the focal adhesion, ECM-receptor and proteoglycan interaction pathways. Using these selected genes, a deep learning neural network architecture was developed to predict response groups. Hyperparameter tuning of the deep learning model dictated a 4 hidden layer architecture with a predictive accuracy of 91.7\%.  This validates the postulate that cell lines with similar gene expression profiles present similar pan-drug chemotherapeutic sensitivity, and it suggests the potential utility of similar combinatorial biomarkers for selection of potent candidate drugs. \vspace{1em}

@@

\subsection*{Pan-cancer therapeutic response cohorts}
To better understand the impact and predictive ability of transcriptomic dysregulation in chemotherapeutic response, a pan-cancer cohort of cell-line and associated therapeutic sensitivity data were obtained from the GDSC database. This database includes 1,110 cell lines from various tumour types, and is thought to represent a relatively comprehensive pan-cancer set. In addition, the acquired dataset contained therapeutic efficacy information in the form of half-maximal inhibitory concentration (IC$_{50}$) values for 251 chemotherapies. These values correspond to the minimal concentration of therapeutic required to induce cell death in 50\% of the cultured cells, with lower values being associated with improved drug efficacy. The data was used to generate a matrix with cell line and accompanying therapeutic information. This dataset was filtered to exclude therapies with less than 80\% of data for all cell lines, followed by the exclusion of cell lines lacking response data for the drugs retained in the first step.

@@

\subsection*{Identification of pan-cancer therapeutic response cohorts}
Cell line therapeutic sensitivity matrices were used to evaluate whether conventional clinical criteria could separate patients into previously-defined chemotherapeutic response groups. These conventional clinical criteria included anatomic location and solid versus non-solid tumour status, as well as broadly applicable molecular markers -- TP53 and KRAS mutation status \cite{colorectal}, \cite{gi}, \cite{lung}, \cite{breast}. Cell-lines were separated into subgroups based on these criteria and plotted to determine whether these criteria effectively clustered response groups.

Following evaluation of existing classifiers, we attempted to create defined cell-line clusters on the basis of the observed chemotherapeutic response of the pan-cancer cell line sample. We developed a Euclidean distance matrix for the retained cell lines based upon their pan-chemotherapy response. This matrix was then used to identify the minimum number of clusters capable of representing the therapeutic heterogeneity identified across the cancer cell lines while maintaining significant inter-cluster distance. K-means clustering was then utilized to assign cell line candidates to appropriate therapeutic response cohorts. Generalized differences in chemotherapeutic efficacy between cohorts were visualized using a heatmap generated by the Pheatmap package in R. Separation between clusters was also visualized using principal component analysis with the factoextra package in R. Following the identification of defined clusters, differences in therapeutic efficacy between the identified cohorts were evaluated. Mann-Whitney U tests were utilized to compare IC$_{50}$ values between the groups. False discovery rate (FDR) correction was utilized to correct for multiple comparisons.

@@

\subsection*{Feature Selection}
To develop a transcriptomic model predictive of therapeutic response clusters, expression data quantified by the GDSC consortium using the Affymetrix U219 microarray for each candidate cell line was obtained. Minimally processed CEL files were obtained from ArrayExpress (ascension number E-MTAB-3610) and processed using the affy package in R. The resulting normalized expression matrix for candidate cell lines was then merged with the existing dataset. This addition resulted in the loss of 7 cell lines (2 from cluster A and 5 from cluster B), resulting in the inclusion of 541 cell-lines in model generation. The microarray dataset quantified the expression of 16,382 genes; a model based on that many features is highly likely to overfit, compromising the generalizability of the model on new data. Such a large feature space also adds unnecessary noise and severely limits the accuracy and computational efficiency of the model which was addressed using dimensionality reduction through feature selection via the BorutaPy package in Python 3  \cite{liu}. The BorutaPy package is a feature selection algorithm based on Random Forest classification which iteratively removes features that are statistically less significant than a shuffled version of the same feature \cite{kursa}. Computational efficiency and the resulting feature set quality were motivators for choosing Boruta over other selection algorithms such as univariate selection or principle component analysis.

@@

To identify defined cohorts of pan-cancer cell lines with similar trends in therapeutic sensitivity, we employed unsupervised clustering of retained cell-lines. Principal component analysis was used to reduce the dimensionality of the dataset, allowing for visualization of defined therapeutic response groups. This process identified two distinct clusters of therapeutic sensitivity (Fig. \ref{fig:pca}), 362 cell lines identified in response group A, and 186 cell lines identified in response group B. The cohorts perform substantially differently in a subset of therapeutics (Fig. \ref{fig:heatmap}). To quantify differences in therapeutic response between clusters, IC50 values were compared between candidate cell lines (Fig. \ref{fig:volcano}). Of the 117 therapies included, 95 had significant differences in efficacy between the two cohorts identified. This suggests that these cohorts represent groups of cell lines with vastly different therapeutic responses. Therefore the ability to accurately stratify into these cohorts may be a valuable tool for stratification prior to chemotherapeutic treatment.

@@

% Discussion
\section{Discussion}
It is well known that there is substantial heterogeneity with respect to chemotherapeutic response between and among cancer types. Although there have been multiple attempts to identify molecular and clinical features predictive of response to particular targeted therapies, there remains considerable variability within subgroups identified using these factors. In this study, we attempt to accurately cluster cancer cell lines into defined groups based on response to a large range of chemotherapeutics, and to create a deep learning transcriptomic model capable of accurately categorizing samples into these defined groups. Using cell-line chemotherapeutic efficacy data obtained from the GDSC consortium, we employ unsupervised clustering techniques to identify two defined therapeutic response groups with significantly different responses to a multitude of standard chemotherapies. We show that these clusters outperform classical clinical criteria for classifying samples into chemotherapeutic response groups, and therefore may prove useful for clinical and research settings. To classify cell lines into these groups, we use a biologically agnostic feature selection algorithm, Boruta, to reduce the original set of 16,382 genes to a subset of 300 genes, which was key to limiting preliminary bias in the model. These genes were then fed into neural networks, which were then optimized. We determined from confusion matrices and ROC curve analysis that the best network architecture utilized 4 hidden layers, and demonstrated a 91.7\% accuracy in classifying response groups. This validates our postulate that gene expression is a prime determinant of chemotherapeutic response, and that cell lines of similar gene expression profile respond similarly to most chemotherapies.

@@

\begin{table*}[!ht]
	\caption{Neural network architectures and the number of neurons per layer ($n=560$).}	\label{tab:architectures}
	\centering
	\begin{tabular}{l l}
		\toprule
		Architecture & Number of neurons \\
		\midrule
		 5 hidden layers & 300 inputs, 2n, 4n, 4n, 2n, n, binary output \\
		 4  hidden layers & 300 inputs, 2n, 4n, 4n, 2n, binary output \\
		 3 hidden layers & 300 inputs, 2n, 4n, 2n, binary output \\
		\bottomrule
	\end{tabular}
\end{table*}

>>>>

% Abstract
\begin{abstract}
	Emerging precision oncology studies have yet to generate a predictive biomarker that utilizes gene expression profiles to stratify tumours into similar pan-drug sensitivity profiles. This development would allow for identification of candidate drugs for treatments that maximize therapeutic response and minimize cytotoxic burden. As such, this study utilized cell line sensitivity and molecular profiling data to generate a combinatorial gene expression predictive biomarker, utilizing feature selection and a deep learning model. A pan-cancer cohort of cell line gene expression data from Genomics of Drug Sensitivity in Cancer (GDSC) was clustered into two response groups. Cell line response groups showed a significant difference in pan-drug chemotherapeutic sensitivity. Due to the high dimensional nature of the microarray data, biologically agnostic feature selection was conducted to identify genes with the highest predictive value. The feature space was reduced to 300 genes, which functional profiling indicated was primarily enriched for the focal adhesion, ECM-receptor and proteoglycan interaction pathways. Using these selected genes, a deep learning neural network architecture was developed to predict response groups. Hyperparameter tuning of the deep learning model dictated a 4 hidden layer architecture with a predictive accuracy of 91.7\%.  This validates the postulate that cell lines with similar gene expression profiles present similar pan-drug chemotherapeutic sensitivity, and it suggests the potential utility of similar combinatorial biomarkers for selection of potent candidate drugs. \vspace{1em}

@@

\subsection*{Pan-cancer therapeutic response cohorts}
To better understand the impact and predictive ability of transcriptomic dysregulation in chemotherapeutic response, a pan-cancer cohort of cell line and associated therapeutic sensitivity data were obtained from the GDSC database. This database includes 1,110 cell lines from various tumour types, and is thought to represent a relatively comprehensive pan-cancer set. In addition, the acquired dataset contained therapeutic efficacy information in the form of half-maximal inhibitory concentration (IC$_{50}$) values for 251 chemotherapies. These values correspond to the minimal concentration of therapeutic required to induce cell death in 50\% of the cultured cells, with lower values being associated with improved drug efficacy. The data was used to generate a matrix with cell line and accompanying therapeutic information. This dataset was filtered to exclude therapies with less than 80\% of data for all cell lines, followed by the exclusion of cell lines lacking response data for the drugs retained in the first step.

@@

\subsection*{Identification of pan-cancer therapeutic response cohorts}
Cell line therapeutic sensitivity matrices were used to evaluate whether conventional clinical criteria could separate patients into previously-defined chemotherapeutic response groups. These conventional clinical criteria included anatomic location and solid versus non-solid tumour status, as well as broadly applicable molecular markers -- TP53 and KRAS mutation status \cite{colorectal, gi, lung, breast}. Cell lines were separated into subgroups based on these criteria and plotted to determine whether these criteria effectively clustered response groups.

Following evaluation of existing classifiers, we attempted to create defined cell line clusters on the basis of the observed chemotherapeutic response of the pan-cancer cell line sample. We developed a Euclidean distance matrix for the retained cell lines based upon their pan-chemotherapy response. This matrix was then used to identify the minimum number of clusters capable of representing the therapeutic heterogeneity identified across the cancer cell lines while maintaining significant inter-cluster distance. K-means clustering was then utilized to assign cell line candidates to appropriate therapeutic response cohorts. Generalized differences in chemotherapeutic efficacy between cohorts were visualized using a heatmap generated by the Pheatmap package in R. Separation between clusters was also visualized using principal component analysis with the factoextra package in R. Following the identification of defined clusters, differences in therapeutic efficacy between the identified cohorts were evaluated. Mann-Whitney U tests were utilized to compare IC$_{50}$ values between the groups. False discovery rate (FDR) correction was utilized to correct for multiple comparisons.

@@

\subsection*{Feature Selection}
To develop a transcriptomic model predictive of therapeutic response clusters, expression data quantified by the GDSC consortium using the Affymetrix U219 microarray for each candidate cell line was obtained. Minimally processed CEL files were obtained from ArrayExpress (ascension number E-MTAB-3610) and processed using the affy package in R. The resulting normalized expression matrix for candidate cell lines was then merged with the existing dataset. This addition resulted in the loss of 7 cell lines (2 from cluster A and 5 from cluster B), resulting in the inclusion of 541 cell lines in model generation. The microarray dataset quantified the expression of 16,382 genes; a model based on that many features is highly likely to overfit, compromising the generalizability of the model on new data. Such a large feature space also adds unnecessary noise and severely limits the accuracy and computational efficiency of the model which was addressed using dimensionality reduction through feature selection via the BorutaPy package in Python 3  \cite{liu}. The BorutaPy package is a feature selection algorithm based on Random Forest classification which iteratively removes features that are statistically less significant than a shuffled version of the same feature \cite{kursa}. Computational efficiency and the resulting feature set quality were motivators for choosing Boruta over other selection algorithms such as univariate selection or principle component analysis.

@@

To identify defined cohorts of pan-cancer cell lines with similar trends in therapeutic sensitivity, we employed unsupervised clustering of retained cell lines. Principal component analysis was used to reduce the dimensionality of the dataset, allowing for visualization of defined therapeutic response groups. This process identified two distinct clusters of therapeutic sensitivity (Fig. \ref{fig:pca}), 362 cell lines identified in response group A, and 186 cell lines identified in response group B. The cohorts perform substantially differently in a subset of therapeutics (Fig. \ref{fig:heatmap}). To quantify differences in therapeutic response between clusters, IC50 values were compared between candidate cell lines (Fig. \ref{fig:volcano}). Of the 117 therapies included, 95 had significant differences in efficacy between the two cohorts identified. This suggests that these cohorts represent groups of cell lines with vastly different therapeutic responses. Therefore the ability to accurately stratify into these cohorts may be a valuable tool for stratification prior to chemotherapeutic treatment.

@@

% Discussion
\section{Discussion}
It is well known that there is substantial heterogeneity with respect to chemotherapeutic response between and among cancer types. Although there have been multiple attempts to identify molecular and clinical features predictive of response to particular targeted therapies, there remains considerable variability within subgroups identified using these factors. In this study, we attempt to accurately cluster cancer cell lines into defined groups based on response to a large range of chemotherapeutics, and to create a deep learning transcriptomic model capable of accurately categorizing samples into these defined groups. Using cell line chemotherapeutic efficacy data obtained from the GDSC consortium, we employ unsupervised clustering techniques to identify two defined therapeutic response groups with significantly different responses to a multitude of standard chemotherapies. We show that these clusters outperform classical clinical criteria for classifying samples into chemotherapeutic response groups, and therefore may prove useful for clinical and research settings. To classify cell lines into these groups, we use a biologically agnostic feature selection algorithm, Boruta, to reduce the original set of 16,382 genes to a subset of 300 genes, which was key to limiting preliminary bias in the model. These genes were then fed into neural networks, which were then optimized. We determined from confusion matrices and ROC curve analysis that the best network architecture utilized 4 hidden layers, and demonstrated a 91.7\% accuracy in classifying response groups. This validates our postulate that gene expression is a prime determinant of chemotherapeutic response, and that cell lines of similar gene expression profile respond similarly to most chemotherapies.

@@

\begin{table*}[!ht]
	\caption{Neural network architectures and the number of neurons per layer ($n=560$).}	\label{tab:architectures}
	\centering
	\begin{tabular}{l l}
		\toprule
		Architecture & Number of neurons \\
		\midrule
		 5 hidden layers & 300 inputs, 2$n$, 4$n$, 4$n$, 2$n$, $n$, binary output \\
		 4  hidden layers & 300 inputs, 2$n$, 4$n$, 4$n$, 2$n$, binary output \\
		 3 hidden layers & 300 inputs, 2$n$, 4$n$, 2$n$, binary output \\
		\bottomrule
	\end{tabular}
\end{table*}
